Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-04-06 src/tex/ms.tex
Update ms.tex
q discussion refinement

\begin{table}
    \caption{Events with bimodality in the $\kappa$ posterior, the \ac{GR} measurement of their detector-frame total mass ($M$), their $\chi_p$ and $\chi_{\rm eff}$ measured without birefringence, and the credible level of $\kappa = 0$ (CL) with birefringence.}
    \begin{ruledtabular}
        \variable{output/bimodal_events_mass.txt}
    \end{ruledtabular}
    \label{tab:bimodal_events_mass}
\end{table}

@@

\citet{Okounkova_2022} gave a constraint on frequency-independent \ac{GW} amplitude birefringence by performing statistical analysis on GWTC-2.
We take $f=100\, \mathrm{Hz}$ as the reference frequency to make the comparison.
They were able to constrain $\tilde{\kappa}$ at $d_C=d_C(z=0.3)$ to be $|\tilde{\kappa}| \lesssim 0.74$ at 1 $\sigma$, where
\begin{equation}
    \tilde{\kappa}d_C = \left[{\kappa}\frac{d_C}{1\, \mathrm{Gpc}}\frac{f}{100\, \mathrm{Hz}}\right]_{f=100\, \mathrm{Hz}}\,.
\end{equation}
Note that $\tilde{\kappa}$ is in units of $1/\mathrm{Gpc}$, while $\kappa$ is dimensionless.
Equivalently, their constraint is $|\kappa| \lesssim 0.74$ at 1 $\sigma$.
We obtained a tighter constraint on $|\kappa|$ with $|\kappa| <$ \variable{output/restricted_absolute_kappa_68.txt}at $68.27\%$ upper limit.

@@

\begin{figure}
    \script{mass_ratio.py}
    \includegraphics[width=\columnwidth]{figures/mass_ratio.pdf}
    \caption{
    Impact of birefringence on mass ratio and precession inference. Posterior on mass ratio $q \equiv m_2/m_1$ (left) and precessing spin $\chi_p$ (right) for GW170104 (top) and GW191105 (bottom), as obtained from both the birefringent analysis (orange) and the regular \ac{GR} analysis (blue).
    }
    \label{fig:mass_ratio}
\end{figure}

@@

As a consequence, a system that would be inferred to be highly spinning in \ac{GR} may instead be inferred to be low-spinning but highly birefringent if nonzero values of $\kappa$ are allowed.
These morphological degeneracies can interact with the distribution of prior probability mass to yield a preference for nonzero $\kappa$ in some events: in absence of a strong constraint from the data, it can be more favorable to place a source at far distances with a large $|\kappa|$, than at close distances with a fine tuned spin configuration.
The resulting posterior can be unimodal or bimodal depending on the specific phase and polarization states observed at the detector.
The events GW170818 (Sec.~\ref{sec:GW170818}) and GW190521 (Sec.~\ref{sec:GW190521}) appear to fall in these two categories respectively.

@@

Besides spins, other parameters can interact with $\kappa$ either directly or through their own couplings with the spins.
Since the mass ratio and spins can affect the phasing of a \ac{GW} inspiral at similar order \cite{Blanchet:2013haa}, we might expect some degree of interaction between $q$ and $\kappa$.
This seems to indeed be the case for some events, like GW170104 or GW191105, as we demonstrate in Fig.~\ref{fig:mass_ratio}.
In the former, introducing birefringence increases the preference for asymmetric masses ($q < 1$), while it decreases the inferred level of precession; in the latter, birefringence increases the preference for symmetric masses ($q = 1$) and does not affect the inference on precession.
In neither case is $\chi_{\rm eff}$ significantly impacted by birefringence.
This degeneracy is likely more limited since mass ratio alone cannot produce significant amplitude modulations.

>>>>

\begin{table}
    \caption{Events with bimodality in the $\kappa$ posterior, the \ac{GR} measurement of their detector-frame total mass ($M$), precessing spin $\chi_p$ and effective spin $\chi_{\rm eff}$, as well as the credible level of $\kappa = 0$ (CL) from the birefringence analysis.}
    \begin{ruledtabular}
        \variable{output/bimodal_events_mass.txt}
    \end{ruledtabular}
    \label{tab:bimodal_events_mass}
\end{table}

@@

\citet{Okounkova_2022} gave a constraint on frequency-independent \ac{GW} amplitude birefringence by performing statistical analysis on GWTC-2.
We take $f=100\, \mathrm{Hz}$ as the reference frequency to make the comparison.
They were able to constrain $\tilde{\kappa}$ at $d_C=d_C(z=0.3)$ to be $|\tilde{\kappa}| \lesssim 0.74$ at 1 $\sigma$, where
\begin{equation}
    \tilde{\kappa}d_C = \left[{\kappa}\frac{d_C}{1\, \mathrm{Gpc}}\frac{f}{100\, \mathrm{Hz}}\right]_{f=100\, \mathrm{Hz}}\,.
\end{equation}
Note that $\tilde{\kappa}$ is in units of $1/\mathrm{Gpc}$, while $\kappa$ is dimensionless.
Equivalently, their constraint is $|\kappa| \lesssim 0.74$ at 1 $\sigma$.
We obtained a tighter $68.27\%$ upper limit of $|\kappa| <$ \variable{output/restricted_absolute_kappa_68.txt}.

@@

\begin{figure}
    \script{mass_ratio.py}
    \includegraphics[width=\columnwidth]{figures/mass_ratio.pdf}
    \caption{
    Posterior on mass ratio $q \equiv m_2/m_1$ (left) and precessing spin $\chi_p$ (right) for GW170104 (top) and GW191105 (bottom), as obtained from both the birefringent analysis (orange) and the regular \ac{GR} analysis (blue).
    }
    \label{fig:mass_ratio}
\end{figure}

@@

As a consequence, a system that would be inferred to be highly spinning in \ac{GR} may instead be inferred to be low-spinning but highly birefringent if nonzero values of $\kappa$ are allowed.
These morphological degeneracies can interact with the distribution of prior probability mass to yield a preference for nonzero $\kappa$ in some events: in absence of a strong constraint from the data, it can be more favorable to place a source at far distances with large $|\kappa|$, than at close distances with a fine tuned spin configuration and low $|\kappa|$.
The resulting posterior can be unimodal or bimodal depending on the specific phase and polarization states observed at the detector.
The events GW170818 (Sec.~\ref{sec:GW170818}) and GW190521 (Sec.~\ref{sec:GW190521}) appear to fall into these two categories respectively.

@@

Besides spins, other parameters can interact with $\kappa$ either directly or through their own couplings with the spins.
Since the mass ratio and can affect the phasing of a \ac{GW} inspiral at similar order as the spins \cite{Blanchet:2013haa}, we might expect some degree of interaction between $q$ and $\kappa$.
This seems to indeed be the case for some events, like GW170104 or GW191105, as we demonstrate in Fig.~\ref{fig:mass_ratio}.
In the former, introducing birefringence increases the preference for asymmetric masses ($q < 1$) while decreasing the inferred level of precession; in the latter, birefringence increases the preference for symmetric masses ($q = 1$) but does not affect the inference on precession.
The $\kappa$ posteriors are bimodal for both events, but do not rule out $\kappa = 0$ (Fig.~\ref{fig:violin_kappa}).
In neither case is $\chi_{\rm eff}$ significantly impacted by birefringence.
The coupling between $q$ and $\kappa$ is likely limited by the fact that changes in the mass ratio alone cannot produce significant amplitude modulations.
